Natural Selection Across Three Generations of Americans

### 2024-02-27 hrs-selection.qmd
Phenotypic change updated in text

To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD. Both of these showed large differences in @hugh2022human, and have significant main effects in this data. Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups. All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA. However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction. [XXX or we just say "nothing survives B. C.]

@@

<!--# REMINDER TO RE-ESTIMATE THE PHENOTYPIC EFFECT WHEN ALL THE CODE HAS BEEN REDONE. -->

To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability. For cognitive performance we have estimated a genetic change of $-0.066$ standard deviations per generation, assuming a twin heritability of $0.51$. This implies a phenotypic change of $-0.066\times0.51^\frac{1}{2}=-0.047$ standard deviations per generation, equivalent to $-0.71$ points in the units of IQ, where a standard deviation is equal to $15$ points. This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype. Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.

@@

# Testing the R2 of the ADHD PGS
rand$variables |>
  with(cor.test(adhd_score, ADHD,, use = "pairwise.complete")) -> adhd_cor

# Writing data for pgi_correct to use
readr::write_tsv(
  rand$variables |> filter(ethnicity == "white"),
  "pgi_correct_data.txt",
  quote = "none"
)

@@

corrected_effects_snp <- corrected_effects |>
    mutate(estimate = estimate*sqrt(chip_h2/R2),
           std.error = std.error*sqrt(chip_h2/R2),
           conf.low = conf.low*sqrt(chip_h2/R2),
           conf.high = conf.high*sqrt(chip_h2/R2),
           heritability = "Chip")

corrected_effects_twin <- corrected_effects |>
    mutate(estimate = estimate*sqrt(twin_h2/R2),
           std.error = std.error*sqrt(twin_h2/R2),
           conf.low = conf.low*sqrt(twin_h2/R2),
           conf.high = conf.high*sqrt(twin_h2/R2),
           heritability = "Twin")

@@

corrected_effects |>
  select(trait, estimate, std.error, Heritability, chip_h2, twin_h2, R2, 
           TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
  readr::write_csv("Corrected_effects.csv")

@@

@@

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is similar to previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $-0.71$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.

>>>>

To complement the above cross-PGS analysis, we focus on two scores which we a priori expect to show intergroup differences: educational attainment and ADHD. Both of these showed large differences in @hugh2022human, and have significant main effects in this data. Figure @fig-top-pgs-interactions displays coefficients of these scores on RLRS for each of the different groups. All differences are in the expected direction, with coefficients being larger for the disadvantaged group, except for low- vs- high-education and EA. However, the only significant difference is between the low- and high-income groups for the coefficient of EA, and this does not survive Bonferroni correction. \[XXX or we just say "nothing survives B. C.\]

@@

@@

@@

corrected_effects_snp <- corrected_effects |>
    mutate(estimate = estimate*sqrt(chip_h2/R2),
           std.error = std.error*sqrt(chip_h2/R2),
           conf.low = conf.low*sqrt(chip_h2/R2),
           conf.high = conf.high*sqrt(chip_h2/R2),
           expected_phenotypic_change = estimate*sqrt(chip_h2),
           heritability = "Chip")

corrected_effects_twin <- corrected_effects |>
    mutate(estimate = estimate*sqrt(twin_h2/R2),
           std.error = std.error*sqrt(twin_h2/R2),
           conf.low = conf.low*sqrt(twin_h2/R2),
           conf.high = conf.high*sqrt(twin_h2/R2),
           expected_phenotypic_change = estimate*sqrt(twin_h2),
           heritability = "Twin")

@@

corrected_effects |>
  select(trait, estimate, std.error, expected_phenotypic_change, Heritability, chip_h2, 
         twin_h2, R2, TwinH2_Source, PoldermanICF_ICD10_subchapter, Category) |> 
  readr::write_csv("Corrected_effects.csv")

####################################
# Extracting Phenotypic change
####################################

IQgen_change <- corrected_effects_twin |> 
                filter(pgs == "CP") |>
                pull(estimate)

IQphen_change_sd_units <- corrected_effects_twin |> 
                filter(pgs == "CP") |>
                pull(expected_phenotypic_change)
IQphen_change <- IQphen_change_sd_units*15

@@

To estimate how natural selection will contribute to changing the phenotype, we can multiply the change in the mean genetic value by the correlation between genetic values and phenotypes $h$, which is the square root of heritability. For cognitive performance we have estimated a genetic change of $`r IQgen_change`$ standard deviations per generation, assuming a twin heritability of $0.51$. This implies a phenotypic change of $`r IQgen_change`\times0.51^\frac{1}{2}=`r IQphen_change_sd_units`$ standard deviations per generation, equivalent to $`r IQphen_change`$ points in the units of IQ, where a standard deviation is equal to $15$ points. This calculation assumes the heritability of the trait remains constant, that the genetic correlation across time is equal to one and it ignores the environmental contributions to changes in the phenotype. Given the assumptions required for this calculation, on top of the problems involved in estimating the genetic change, it can only be considered a guess.

@@

Lastly, we provide the first estimates of selection differentials of "true" polygenic scores for traits other than intelligence and education, finding results typically between 0-0.2 standard deviations. Our estimate of how much natural selection is changing intelligence is similar to previous work. For example, @kong2017selection provide an estimate of $-0.9$ IQ points every $30$ years in Iceland, compared to our $`r round(IQphen_change,1)`$ points per generation. However, cognitive traits are only one part of the story of natural selection in humans. Whilst past research has focused on these traits, we find selection differentials are often of similar magnitudes for health related traits. The selection differential for self-rated health is greater than those for education and cognitive performance. The most significant, positively selected trait was ADHD, which was also found in the UK [@hugh2022human]. Future research should study the health and medical implications of natural selection, in addition to their social implications. To know more, we must await more accurate polygenic scores.
